Dynamic Load Balancing for\\Heterogeneous Systems

### 2011-12-05 paper/paper.tex
Spelling fixes.

In addition, the CPU typically has a deep pipeline to increase throughput of the system.
Longer and longer pipelines increase the utilization of every element found on the CPU,
further increasing latency-critical performance.  Performance of this pipeline is in danger
of being flushed on braching codes, but advanced branch prediction and speculative execution
are able to greatly reduce the amount of times where a pipeline flush must occur.  Out-of-order
execution of instructions in the program also improves performance for single threads on the
CPU.

@@

Each processor on the GPU is very dumb and primitive compared to the processors on the CPU.
There is no speculative execution, deep pipelines, out-of-order exeuction, or branch prediction
on this device.  Instead, the space saved by not having these features is instead used to
cram more and more compute cores onto the device.  This strategy of having a large number of
very simple cores goes back to the orginial architectual decisions for the GPU.  Graphics applications
are relatively straight-forward and require a lot of math and very little complex coding.
Therefore, for graphics, there is little performance lost for not having these features, whereas
for GPGPU applications these features can greatly diminish overall kernel performance.

Groups of processing units, Streaming Multiprocessors (SMs), share a single instruction
dispatch unit and operate in lock step.  The threads themselves run on each SM
in a group, called a warp, typically of 32 threads.  Because of these architectural
features divergent branching in a warp (caused by if-else clauses) greatly degrade application
performance.  Unlike the CPU, the GPU is unable to avoid warp-level divergence of
the different threads, and therefore must always take the penalty of braching.

@@

One of the main benefits of the APU is the elimination of the PCI-e
bus, which allows for much faster transfer between CPU and GPU.  Daga et al.
show that the AMD Fusion architecture allows for much faster data transfers
between CPU and GPU, allowing for an overall improvement in runtime, despite
reduced compute power on the GPU~\cite{Daga2011}.  The improvment of the 
data transfer times greatly improves the applicability of the GPU for use
in general computing paradigms, specifically that of CPU+GPU co-processing.
Decreasing the data trasnsfer time reduces the penalty associated with migrating
computation to the GPU, and therefore can access the computation of the GPU
more at will then in discrete GPUs which are bound by the slow access speeds
of the PCI-e bus.

@@

The next benchmark, referred henceforth as the ``Vector Add+'' benchmark, is similar to
the VectorAdd benchmark, but it computes 1000 additions before storing the result
in the output vector.  The motivation behind this benchmark is to reduce memory
access overhead with more computation.  This benchmark was compiled without optimizations
so that the additions would not be optmized into a single multiplication.  The data
dependencies are exactly the same as for the VectorAdd application, so the results for
data transfer speeds should be equal for both applications, and the difference between
the two should only be the kernel execution times.

@@

\section{The OpenCL Framework}
For this project, we will use the OpenCL Framework for our kernel lauching.  OpenCL
is an application framework for heterogeneous parallel computing.  It includes an
API for creating and transferring memory to accleration devices, as well as dynamic
compilation of application kernels that are run on the acceleators.Using OpenCL 
allows us to write the host code only once, while allowing us to run the core 
application kernel on the CPU, GPU, or both.

In OpenCL each thread is logically mapped to a work-item.  These work-items are grouped into
workgroups.  These workgroups map to a SM on a discrete GPU.  For instance, these workgroups
share access to a SMs local memory and can perform a barrier with all the other work-items
in the workgroup.  In the case of a CPU acclerator, the workgroups still exist, but optimizations
such as manual caching via local memory will not be effective.

@@

For our purposes, we decided to use a seperate context for each device, as well as a seperate
command queue for each context.  We decided to use a seperate context for each device, as it would
allow for finer-grained control of data transfers and residency. Using a seperate command queue
for each device will facilitate multiple devices to be used simultaneously, as each command queue
operates on its own seperate thread.

@@

The first load balancing scheme that was implemented was a static division of work between
the CPU and the GPU at a specified ratio.  This ratio was user-defined allowing for best 
performance after runtime benchmarking (i.e. the precentage of work done on the CPU and GPU).
The cases of CPU-only and GPU-only scheduling are just special cases of this scheduler, of
the ratios 0 and 1 respectively.

@@

To address the issue of a dynamic work load between workgroups for kernels, we also used
a dynamic load-balancing scheme.  In this scheme the workgroups were grouped into larger
chunks.  Each device is initially allocated one chunk of work to work on, and as each chunk
is retired by the device a new chunk is allocated.  Each device is given a seperate hardware
thread, and the thread will wait until the current chunk is finished, then it will synchronize
through a shared variable to acquire the next chunk of work for the device.  This process
will continue until the work queue is out of work.

@@

First, the reduction benchmark was run on the discrete GPU platform with a variety of
different data sizes and load balancing policies, as shown in Figure \ref{fig:reduce_discrete}.
Interestingly, the CPU outperformed the GPU on this particular benchmark.  This is likely
due to the GPU not utilizing all of its stream processors during a reduction operation, as
discussed previously.  It is also the case that the large amount of synchronization that must
occur in each iteration of the GPU reduction, and the number of kernel launches required for
each iteration that is not present in the CPU reduction. In addition, because the CPU had better
performance on the kernel alone, it saw an even greater improvement when the data transfer costs
were considered, with the CPU-only case requiring no data movement at all. Unfortunately, none of
the static load balancing schemes were able to outperform the CPU-only scheme; all the static times were
somewhere in between the CPU-only and GPU-only times.  Additionally, the dynamic load balancing
scheme performed significantly worse than all the other schemes, and more than 100\%
worse than just the CPU.  The cause for this degredation in performance was likely due to the
overhead of scheduling threads and communciation between those threads, as well as moving data to
the GPU in small chunks when the GPU showed very poor performance compared to the CPU.

@@

The results for the VectorAdd+ benchmark on the discrete GPU platform are shown in Figure
\ref{fig:vector_plus_discrete}.  There are a few notable trends in this figure that are not
present in the previous two.  The first is is how drastically the GPU outperforms the CPU --
with over a 100x speedup over just the CPU with 16~MB of data.  Because of the very large
amount of computation that needs to be performed, compared to the low amount of data accesses
required in the kernel (3 data accesses for 1000 additions), the GPU is able to outperform
the CPU because of its very large throughput for computationally intensive kernels.  The second 
thing to notice for this benchmark specificially is that the
dynamic load balancing drastically outperforms the static load balancing schemes.  This is
almost certainly because the GPU so drastically outperforms the CPU.  With the dynamic load
balancing scheme, only small chunks of data are given to the CPU at a time, rather than a
static percentage, allowing for the GPU to take advantage of much more of its speedup over
the CPU. Increasing the chunk size for the GPU (taking two chunks instead of one at a time)
would likely yield in a greater speedup for the dynamic load-balancing case.

@@

The above results have lead us to believe that the CPU and GPU kernels are
not running truly in parallel as we initially believed.  As such, we used
the OpenCL profiling capabilities to report how long each of the kernels took using the static
load balancer with 95\% of the data on the CPU.  Figure \ref{fig:sequential_discrepancy}
shows the discrepancy between the actual total runtime of the vector add
benchmark and the expected runtime of sequential calls to the kernels 
(that is, the sum of the CPU kernel runtime
and the GPU kernel runtime).  If the kernels were truly running in parallel,
we would expect to see a negative discrepancy between the actual runtime and
expected runtime.  It was impossible for us to simply time the start and end
times of the different kernels, because while that information was available
for each kernel launch, it was kept as device-speciifc times. We could not
reconcile the two different timing schemes, because of the different clock
frequencies and the lack of a mapping from time from clock start for either
device.

>>>>

In addition, the CPU typically has a deep pipeline to increase throughput of the system.
Longer and longer pipelines increase the utilization of every element found on the CPU,
further increasing latency-critical performance.  Performance of this pipeline is in danger
of being flushed on branching codes, but advanced branch prediction and speculative execution
are able to greatly reduce the amount of times where a pipeline flush must occur.  Out-of-order
execution of instructions in the program also improves performance for single threads on the
CPU.

@@

Each processor on the GPU is very dumb and primitive compared to the processors on the CPU.
There is no speculative execution, deep pipelines, out-of-order execution, or branch prediction
on this device.  Instead, the space saved by not having these features is instead used to
cram more and more compute cores onto the device.  This strategy of having a large number of
very simple cores goes back to the original architectural decisions for the GPU.  Graphics applications
are relatively straight-forward and require a lot of math and very little complex coding.
Therefore, for graphics, there is little performance lost for not having these features, whereas
for GPGPU applications these features can greatly diminish overall kernel performance.

Groups of processing units, Streaming Multiprocessors (SMs), share a single instruction
dispatch unit and operate in lock step.  The threads themselves run on each SM
in a group, called a warp, typically of 32 threads.  Because of these architectural
features divergent branching in a warp (caused by if-else clauses) greatly degrade application
performance.  Unlike the CPU, the GPU is unable to avoid warp-level divergence of
the different threads, and therefore must always take the penalty of branching.

@@

One of the main benefits of the APU is the elimination of the PCI-e
bus, which allows for much faster transfer between CPU and GPU.  Daga et al.
show that the AMD Fusion architecture allows for much faster data transfers
between CPU and GPU, allowing for an overall improvement in runtime, despite
reduced compute power on the GPU~\cite{Daga2011}.  The improvement of the 
data transfer times greatly improves the applicability of the GPU for use
in general computing paradigms, specifically that of CPU+GPU co-processing.
Decreasing the data transfer time reduces the penalty associated with migrating
computation to the GPU, and therefore can access the computation of the GPU
more at will then in discrete GPUs which are bound by the slow access speeds
of the PCI-e bus.

@@

The next benchmark, referred henceforth as the ``Vector Add+'' benchmark, is similar to
the VectorAdd benchmark, but it computes 1000 additions before storing the result
in the output vector.  The motivation behind this benchmark is to reduce memory
access overhead with more computation.  This benchmark was compiled without optimizations
so that the additions would not be optimized into a single multiplication.  The data
dependencies are exactly the same as for the VectorAdd application, so the results for
data transfer speeds should be equal for both applications, and the difference between
the two should only be the kernel execution times.

@@

\section{The OpenCL Framework}
For this project, we will use the OpenCL Framework for our kernel launching.  OpenCL
is an application framework for heterogeneous parallel computing.  It includes an
API for creating and transferring memory to acceleration devices, as well as dynamic
compilation of application kernels that are run on the accelerators.Using OpenCL 
allows us to write the host code only once, while allowing us to run the core 
application kernel on the CPU, GPU, or both.

In OpenCL each thread is logically mapped to a work-item.  These work-items are grouped into
workgroups.  These workgroups map to a SM on a discrete GPU.  For instance, these workgroups
share access to a SMs local memory and can perform a barrier with all the other work-items
in the workgroup.  In the case of a CPU accelerator, the workgroups still exist, but optimizations
such as manual caching via local memory will not be effective.

@@

For our purposes, we decided to use a separate context for each device, as well as a separate
command queue for each context.  We decided to use a separate context for each device, as it would
allow for finer-grained control of data transfers and residency. Using a separate command queue
for each device will facilitate multiple devices to be used simultaneously, as each command queue
operates on its own separate thread.

@@

The first load balancing scheme that was implemented was a static division of work between
the CPU and the GPU at a specified ratio.  This ratio was user-defined allowing for best 
performance after runtime benchmarking (i.e. the percentage of work done on the CPU and GPU).
The cases of CPU-only and GPU-only scheduling are just special cases of this scheduler, of
the ratios 0 and 1 respectively.

@@

To address the issue of a dynamic work load between workgroups for kernels, we also used
a dynamic load-balancing scheme.  In this scheme the workgroups were grouped into larger
chunks.  Each device is initially allocated one chunk of work to work on, and as each chunk
is retired by the device a new chunk is allocated.  Each device is given a separate hardware
thread, and the thread will wait until the current chunk is finished, then it will synchronize
through a shared variable to acquire the next chunk of work for the device.  This process
will continue until the work queue is out of work.

@@

First, the reduction benchmark was run on the discrete GPU platform with a variety of
different data sizes and load balancing policies, as shown in Figure \ref{fig:reduce_discrete}.
Interestingly, the CPU outperformed the GPU on this particular benchmark.  This is likely
due to the GPU not utilizing all of its stream processors during a reduction operation, as
discussed previously.  It is also the case that the large amount of synchronization that must
occur in each iteration of the GPU reduction, and the number of kernel launches required for
each iteration that is not present in the CPU reduction. In addition, because the CPU had better
performance on the kernel alone, it saw an even greater improvement when the data transfer costs
were considered, with the CPU-only case requiring no data movement at all. Unfortunately, none of
the static load balancing schemes were able to outperform the CPU-only scheme; all the static times were
somewhere in between the CPU-only and GPU-only times.  Additionally, the dynamic load balancing
scheme performed significantly worse than all the other schemes, and more than 100\%
worse than just the CPU.  The cause for this degradation in performance was likely due to the
overhead of scheduling threads and communication between those threads, as well as moving data to
the GPU in small chunks when the GPU showed very poor performance compared to the CPU.

@@

The results for the VectorAdd+ benchmark on the discrete GPU platform are shown in Figure
\ref{fig:vector_plus_discrete}.  There are a few notable trends in this figure that are not
present in the previous two.  The first is is how drastically the GPU outperforms the CPU --
with over a 100x speedup over just the CPU with 16~MB of data.  Because of the very large
amount of computation that needs to be performed, compared to the low amount of data accesses
required in the kernel (3 data accesses for 1000 additions), the GPU is able to outperform
the CPU because of its very large throughput for computationally intensive kernels.  The second 
thing to notice for this benchmark specifically is that the
dynamic load balancing drastically outperforms the static load balancing schemes.  This is
almost certainly because the GPU so drastically outperforms the CPU.  With the dynamic load
balancing scheme, only small chunks of data are given to the CPU at a time, rather than a
static percentage, allowing for the GPU to take advantage of much more of its speedup over
the CPU. Increasing the chunk size for the GPU (taking two chunks instead of one at a time)
would likely yield in a greater speedup for the dynamic load-balancing case.

@@

The above results have lead us to believe that the CPU and GPU kernels are
not running truly in parallel as we initially believed.  As such, we used
the OpenCL profiling capabilities to report how long each of the kernels took using the static
load balancer with 95\% of the data on the CPU.  Figure \ref{fig:sequential_discrepancy}
shows the discrepancy between the actual total runtime of the vector add
benchmark and the expected runtime of sequential calls to the kernels 
(that is, the sum of the CPU kernel runtime
and the GPU kernel runtime).  If the kernels were truly running in parallel,
we would expect to see a negative discrepancy between the actual runtime and
expected runtime.  It was impossible for us to simply time the start and end
times of the different kernels, because while that information was available
for each kernel launch, it was kept as device-specific times. We could not
reconcile the two different timing schemes, because of the different clock
frequencies and the lack of a mapping from time from clock start for either
device.
